---
title: "Project EDA"
author: "Lilly Clark, Thea Dowrich, Daisy Fang"
date: "04-10-20"
output: 
  html_document:
    #code_folding: hide
    #keep_md: yes
    theme: spacelab
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages}
library(tidyverse)
library(tidyverse)
library(maps)
library(viridis)
library(patchwork)
```

```{r load-data}
facilities <- read.csv("data/all_health_facilities.csv")
insurance <- read.csv("data/insurance.csv")
inpatients <- read_csv("data/medicare_inpatients.csv")
medicare <- read_csv("data/medicare.csv")
```

## Inpatient Data 
From data.cms.gov: The Inpatient Utilization and Payment Public Use File (Inpatient PUF) provides information on inpatient discharges for Medicare fee-for-service beneficiaries. The Inpatient PUF includes information on utilization, payment (total payment and Medicare payment), and hospital-specific charges for the more than 3,000 U.S. hospitals that receive Medicare Inpatient Prospective Payment System (IPPS) payments. The PUF is organized by hospital and Medicare Severity Diagnosis Related Group (MS-DRG) and covers Fiscal Year (FY) 2017. MS-DRGs included in the PUF represent more than 7 million discharges or 75 percent of total Medicare IPPS discharges.

```{r}
inpatients <- inpatients %>%
  mutate(state = as.factor(state),
         drg = as.factor(str_extract(drg_definition, "\\d{3}")),
         drg_definition = as.factor(str_remove(drg_definition, "\\d{3} - ")),
         average_outofpocket = average_total_payments - average_medicare_payments,
         city = as.factor(city))
```

```{r}
medi <- left_join(inpatients, medicare, by = "name")
medi_filt <- medi %>%
  filter(!is.na(id.y))
```

```{r}
cities <- medi_filt %>%
  group_by(city.x, ) %>%
  summarize(mean = mean(average_outofpocket)) %>%
  arrange(desc(mean))

cities_loc <- medi_filt %>%
  select(city.x, state.x, long, lat)

cities <- left_join(cities, cities_loc, by = "city.x") %>%
  mutate(mean = round(mean, 2))
```

```{r}
inpatients %>%
  group_by(drg_definition) %>%
  summarize(mean_outofpocket_costs = mean(average_outofpocket)) %>%
  arrange(desc(mean_outofpocket_costs)) %>%
  head(10)
```

```{r}
medicare_ratings <- medicare %>%
  mutate(
    effectiveness_of_care_national_comparison = case_when(
      effectiveness_of_care_national_comparison == "Not Available" ~ as.character(NA),
      TRUE ~ effectiveness_of_care_national_comparison),
    timeliness_of_care_national_comparison = case_when(
      timeliness_of_care_national_comparison == "Not Available" ~ as.character(NA),
      TRUE ~ timeliness_of_care_national_comparison),
    timeliness_of_care_national_comparison = as.factor(timeliness_of_care_national_comparison),
    effectiveness_of_care_national_comparison = as.factor(effectiveness_of_care_national_comparison),
    effectiveness_of_care_national_comparison = fct_relevel(effectiveness_of_care_national_comparison,
                                                            c("Below the national average", 
                                                              "Same as the national average", 
                                                              "Above the national average")),
    timeliness_of_care_national_comparison = fct_relevel(timeliness_of_care_national_comparison,
                                                            c("Below the national average", 
                                                              "Same as the national average", 
                                                              "Above the national average"))
  )
```

```{r}
us_map <- map_data("state")
```

```{r}
cols <- c("Below the national average" = "red", "Above the national average" = "#0A97F0", "Same as the national average" = "black")
```

```{r}
medicare_locations_plot <- medicare %>%
  filter(state != "HI", state != "AK", long > -140, lat > 22) %>%
  ggplot() +
  geom_polygon(data = us_map, aes(x=long, y=lat, group=group),
                   color="#9EA5A9", fill = "#CED5DA") +
  geom_point(aes(x = long, y = lat), alpha = .2) +
  coord_map() +
  theme_void() +
  theme(plot.caption = element_text(hjust = .5)) +
  labs(title = "Locations of Medicare Providers", subtitle = "Medicare providers are spread thin in some Midwestern states.")
```

```{r}
medicare_ratings_map1 <- medicare_ratings %>%
    filter(state != "HI", state != "AK", long > -140, lat > 22,
         !is.na(effectiveness_of_care_national_comparison))

effectiveness_plot <- medicare_ratings_map1 %>%
  ggplot() +
  geom_polygon(data = us_map, aes(x=long, y=lat, group=group),
                   color="#9EA5A9", fill = "#CED5DA") +
  geom_point(data = subset(medicare_ratings_map1, 
                           effectiveness_of_care_national_comparison == "Same as the national average"),
             aes(x = long, y = lat, color = effectiveness_of_care_national_comparison), 
             alpha = .2) +
  geom_point(data = subset(medicare_ratings_map1, 
                           effectiveness_of_care_national_comparison == "Above the national average"),
             aes(x = long, y = lat, color = effectiveness_of_care_national_comparison), 
             alpha = .6) +
  geom_point(data = subset(medicare_ratings_map1, 
                           effectiveness_of_care_national_comparison == "Below the national average"),
             aes(x = long, y = lat, color = effectiveness_of_care_national_comparison), 
             alpha = .4) +
  scale_color_manual(values = cols) +
  coord_map() +
  theme_void() +
  theme(plot.caption = element_text(hjust = .5)) +
  labs(title = "National Comparison of Effectiveness of Care for Medicare Providers", 
       subtitle = "Apparent clusters of ineffective providers in New York and Chicago areas.",
       color = "Effectiveness of care")
```

```{r}
medicare_ratings_map2 <- medicare_ratings %>%
    filter(state != "HI", state != "AK", long > -140, lat > 22,
         !is.na(timeliness_of_care_national_comparison))

timeliness_plot <- medicare_ratings_map2 %>%
  ggplot() +
  geom_polygon(data = us_map, aes(x=long, y=lat, group=group),
                   color="#9EA5A9", fill = "#CED5DA") +
  geom_point(data = subset(medicare_ratings_map2, 
                           timeliness_of_care_national_comparison == "Same as the national average"),
             aes(x = long, y = lat, color = timeliness_of_care_national_comparison), 
             alpha = .2) +
  geom_point(data = subset(medicare_ratings_map2, 
                           timeliness_of_care_national_comparison == "Above the national average"),
             aes(x = long, y = lat, color = timeliness_of_care_national_comparison), 
             alpha = .6) +
  geom_point(data = subset(medicare_ratings_map2, 
                           timeliness_of_care_national_comparison == "Below the national average"),
             aes(x = long, y = lat, color = timeliness_of_care_national_comparison), 
             alpha = .4) +
  scale_color_manual(values = cols) +
  coord_map() +
  theme_void() +
  theme(plot.caption = element_text(hjust = .5)) +
  labs(title = "National Comparison of Timeliness of Care for Medicare Providers", 
       subtitle = "Timeliness suffers in East Coast and California cities.",
       color = "Timeliness of care")
```

```{r fig.width = 8, fig.height = 12}
medicare_locations_plot / effectiveness_plot / timeliness_plot +
  plot_layout(widths = 5, heights = 20)
```

```{r fig.width = 8, fig.height = 6}
cities %>%
  arrange(mean) %>%
  filter(state.x != "HI", state.x != "AK", long > -140) %>%
  ggplot() +
  geom_polygon(data = us_map, aes(x=long, y=lat, group=group),
                   color="#9EA5A9", fill = "#CED5DA") +
  geom_point(aes(x = long, y = lat, color = mean, size = mean), alpha = .2) +
  coord_map() +
  scale_size_continuous(range=c(.1,5), guide = FALSE) +
  scale_color_viridis(trans="log") +
  theme_void() +
  theme(plot.caption = element_text(hjust = .5)) +
  labs(title = "Costs Not Covered By Medicare for Inpatient Procedures", color = "Mean out-of-pocket costs ($)", subtitle = "in 2017", caption = "Average personal costs for procedures signicantly higher in some California and East Coast cities.")
```







## Insurance
```{r}
insurance$uninsured_rate_2010 <- as.numeric(str_replace_all(insurance$uninsured_rate_2010, "%", ""))

insurance$uninsured_rate_2015 <- as.numeric(str_replace_all(insurance$uninsured_rate_2015, "%", ""))


ggplot(data = insurance, mapping = aes(x = uninsured_rate_2010, y = uninsured_rate_2015)) +
  geom_point() +
  labs(x = "Insurance Rate in 2010", y = "Insurance Rate in 2015", 
       title = "Changes in Health Insurance Rates before and after the Affordable Care Act")

medicaid_clean <- insurance %>%
  filter(state_medicaid_expansion_2016 == TRUE | state_medicaid_expansion_2016 == FALSE)

ggplot(data = medicaid_clean, mapping = aes(x = state_medicaid_expansion_2016, y = medicaid_enrollment_change_2013_2016)) +
  geom_boxplot() +
  ylim(-4000, 4500000) +
  labs(x = "Did the state expand funding for Medicaid in 2016?", y = "Medicaid Enrollment Change between 2013 and 2016")
  
```

